Iterative Monomorphisation

### 2024-03-22 iterative_monomorphisation.tex
more edits

The iterative monomorphisation algorithm (Figure~\ref{iter_mono}) has two phases. The first phase applies a \emph{monomorphisation step} to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached. This limit is the only bound necessary for the algorithm to terminate. We use the colour blue to identify bounds and code related to bounds. At the end of each of these iterations, the old and new type argument maps are updated with newly generated types. Once these iterations are completed, the first phase is complete and the substitutions used to create new type arguments are passed to the formula generation function for the second phase.

@@

Formula monomorphisation (Figure~\ref{mono_step}) computes new type arguments for a single formula. It forms the core of the monomorphisation process. A set of substitutions obtained from matching type argument tuples is iterated over. The separation of type arguments into old and new maps is put to use here as \emph{matches} (Figure~\ref{subst_gen}) is not called on type argument maps which are both old. This avoids re-computing some of the matchings processed in previous iterations. Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings. Each substitution is applied to all polymorphic type argument tuples such that at least one type variable in the tuple is instantiated by the substitution.

The total number of type arguments can increase cubicly in the number of type arguments at each iteration and can therefore grow doubly exponentially in the number of iterations. We sketch an example of such growth. If we assume that there are after \(k\) iterations \(N_k\) total type arguments split evenly between monomorphic and non-monomorphic type arguments, all in a single formula, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions. Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples. If half of these new type arguments are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = N_k^3 \cdot 2^{-3}\) evenly split type argument tuples. Therefore, the total number of type arguments on the \(k\)th iteration can reach \(N_k = N_0^{3^k} \cdot 2^{\frac{-3^{k+2}+3}{2}}\).

@@

The substitutions used for generating new type arguments are obtained thanks to the \emph{matches} function (Figure~\ref{subst_gen}).
Each symbol instances from \(N(\varphi)\) is matched against all symbol instances from \(M\) of the same symbol.
For two such symbol instances, types from the non-monomorphic type argument tuple are matched against the corresponding types from the monomorphic type argument tuple. The resulting substitutions are composed if they are compatible. In the pseudocode, this is checked by making sure that the \textbf{while} loop has iterated over all elements of the type argument tuple. If any substitutions are incompatible, the matchings are discarded. Since the composition of two compatible substitutions is commutative, the order of composition is irrelevant.

@@

The various bounds presented here overlap to some extent. For instance, limiting the number of substitutions generated at this step may be sufficient to curb the number of new type argument tuples, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous. Each bound nonetheless has its use cases. For example, problems that lead to few successful matches but many type arguments may benefit from a limit on the number of new type arguments whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitution.

@@

Once all monomorphisation iterations have been completed, we are left with a set of substitutions. The last phase uses these substitutions to instantiate the type variables in the input problem's polymorphic formuale. The ordering of the set \(S\) of substitutions can have a significant impact on the formulae resulting from the last phase. As discussed in parameter optimisation \ref{param_opti}, the best results were obtained when substitutions generated in earlier iterations are ordered first.

To monomorphise a polymorphic formula, we first compute its \emph{monomorphising substitutions} (Figure~\ref{mono_substs}). A substitution \(\sigma\) is a monomorphising substitution for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic. Such substitutions are computing using a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S\), it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S\) which instantiates at least one of the type variables in \(V\). It is important that \(\sigma_{\negvthinspace\Delta}\) be compatible with \(\sigma\) so that they may be composed and the function recursively called to instantiate the remaining type variables.

The Zipperposition implementation of this function makes use of a map from type variables to substitutions in place of a set to filter the relevant substitutions from \(S^\text{all}\) more efficiently. The \textcolor{ourblueviolet}{max\_substs} exists for two main reasons:
\begin{enumerate}
   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic clause per initial polymorphic clause. Generating an excessive number of new clauses can lead to the downstream prover being flooded.
   \item This function is the most explosive part of the algorithm. If \(S\) contains \(n\) substitutions which each instantiate exactly one of \(v\) different type variable, up to \(n^v\) monomorphising substititions may be generated. This should be considered in light of the fact that the total number of type argument tuples, which  re used to generate \(S\), is doubly exponential in the number of loop iterations.
\end{enumerate}
%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S_{\text{res}}\) would represent the leaves that have already been visited and \(\sigma\) would represent the current node of the tree.]

>>>>

The iterative monomorphisation algorithm is given in Figure~\ref{iter_mono}. It has two phases. The first phase applies a \emph{monomorphisation step} to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached. This limit is the only bound necessary for the algorithm to terminate. We use the colour blue to identify bounds and code related to bounds. At the end of each of these iterations, the old and new type argument maps are updated with newly generated types. Once these iterations are completed, the first phase is done and the substitutions used to create new type arguments are passed to $\mathit{generate\_mono\_formulae}$ for the second phase.

@@

The formula monomorphisation algorithm is given in Figure~\ref{mono_step}. It forms the core of the monomorphisation process. It computes new type arguments for a single formula. A set of substitutions obtained from matching type argument tuples is iterated over. 
The separation of type arguments into old and new maps is put to use to ensure that only combinations involving at least one new map are considered. This avoids re-computing some of the matchings processed in previous iterations. Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings. Each substitution is applied to all polymorphic type argument tuples such that at least one type variable in the tuple is instantiated by the substitution.

The total number of type argument tuples can increase cubicly in the number of type arguments at each iteration and can therefore grow doubly exponentially in the number of iterations. We sketch an example of such growth. If we assume that there are after \(k\) iterations \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type arguments, all in a single formula, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions. Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples. If half of these new type argument tuples are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = N_k^3 \cdot 2^{-3}\) evenly split type argument tuples. Therefore, the total number of type arguments on the \(k\)th iteration can reach \(N_k = N_0^{3^k} \cdot 2^{\frac{-3^{k+2}+3}{2}}\).

@@

The \emph{matches} function, which computes the substitutions used for generating new type arguments, is given in Figure~\ref{subst_gen}. Each symbol instance from \(N(\varphi)\) is matched against all corresponding symbol instances from \(M\).
For two such symbol instances, types from the non-monomorphic type argument tuple are matched component-wise against the types from the monomorphic type argument tuple. The resulting substitutions are composed if they are compatible. In the pseudocode, this is checked by making sure that the \textbf{while} loop has successfully iterated over all elements of the type argument tuple. If any substitutions are incompatible, the matchings are discarded. Since the composition of two compatible substitutions is commutative, the order of composition is irrelevant.

@@

The various bounds presented here overlap to some extent. For instance, limiting the number of substitutions generated at this step may be sufficient to curb the number of new type argument tuples, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous. Each bound nonetheless has its uses. For example, problems that lead to few successful matches but many type arguments may benefit from a limit on the number of new type arguments whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitution.

@@

Once all monomorphisation iterations have been completed, we are left with a set of substitutions. The last phase uses these substitutions to instantiate the type variables in the input problem's polymorphic formuale. In the presence of bounds, the order in which the elements of the set \(S\) of substitutions is traversed affects the formulae resulting from the last phase. As we will discuss in Section~\ref{param_opti}, the best results were obtained when substitutions generated in earlier iterations are ordered first.

To monomorphise a polymorphic formula, we first compute its {monomorphising substitutions} using the \emph{mono\_substs} function given in Figure~\ref{mono_substs}.
A substitution \(\sigma\) is \emph{monomorphising} for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic. Such substitutions are computed using a recursive function. Given a set \(V\) of type variables and a set \(S\) of substitutions, it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S\) that instantiates at least one of the type variables in \(V\). It is important that \(\sigma_{\negvthinspace\Delta}\) be compatible with \(\sigma\) so that they can be composed and the function recursively called to instantiate the remaining type variables.

The Zipperposition implementation of the \emph{mono\_substs} function uses a map from type variables to substitutions instead of a set to filter the relevant substitutions from \(S^\text{all}\) more efficiently. The \textcolor{ourblueviolet}{max\_substs} bound exists for two main reasons:
\begin{enumerate}
   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic clause per initial polymorphic clause. Generating an excessive number of new clauses can flood the downstream prover.
   \item The \emph{mono\_substs} function is the most explosive part of the algorithm. If \(S\) contains \(n\) substitutions that each instantiate exactly one of \(v\) different type variables, up to \(n^v\) monomorphising substititions may be generated. Recall that the total number of type argument tuples that are used to generate \(S\) is doubly exponential in the number of loop iterations.
\end{enumerate}
%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S_{\text{res}}\) would represent the leaves that have already been visited and \(\sigma\) would represent the current node of the tree.]
